---
title: "Lab 03"
author: "Lakhvinder Singh"
date: "03.24.2026"
output: 
   pdf_document: 
      keep_tex: true
my_subtitle: "Env Analysis in R"
fancy: true
geometry: margin=1in
latex_engine: pdflatex
colorlinks: true
---

# Lab 03: Spatial autocorrelation, globally and locally

### Read the instructions COMPLETELY before starting the lab

This lab builds on many of the discussions and exercises from class, including previous labs.

### Attribution

This lab uses some code examples and directions from https://mgimond.github.io/Spatial/spatial-autocorrelation-in-r.html

### Formatting your submission

This lab must be placed into a public repository on GitHub (www.github.com). Before the due date, submit **on Canvas** a link to the repository. I will then download your repositories and run your code. The code must be contained in either a .R script or a .Rmd markdown document. As I need to run your code, any data you use in the lab must be referenced using **relative path names**. Finally, answers to questions I pose in this document must also be in the repository at the time you submit your link to Canvas. They can be in a separate text file, or if you decide to use an RMarkdown document, you can answer them directly in the doc.

## Data

The data for this lab can be found on the US Census website. 

1. First, go here: https://www.census.gov/geographies/mapping-files/2020/geo/tiger-data.html

2. Second, scroll to the "Download Legal and Administrative Areas Geodatabases" section

3. Click on "County" to download the county data for all of the US (the direct link is also here: https://www2.census.gov/geo/tiger/TIGER_DP/2020ACS/ACS_2020_5YR_COUNTY.gdb.zip). Note, the data are approximately 200 MB in size, so be aware of any connection or storage issues prior to downloading the file.


## Introduction

In this lab, we will be calculating the spatial autocorrelation of various Census variables across a subset of the US. Please note, the dataset you downloaded above is larger than the current 100MB limit GitHub imposes on single files. This means you'll be unable to push that dataset to GitHub. Accordingly, I *strongly* suggest you subset the data such that your files are under this limit. This will be vital when I grade your submissions. If you're not certain how to save a subset of the file to disk, look at ```?sf::write_sf``` for help. We will also be using a new package called ```spdep``` in this assignment.


We begin by loading the relevant packages and data

```{r packages, echo=TRUE, message = TRUE}
library(spdep)
library(sf)
library(tidyverse)
library(tmap)
library(tidycensus)
```


Next, we load our data, look at it, then maybe plot it (the plot might take some time). This file is a geodatabase, a proprietary file format created by ESRI. Conveniently, `sf_read` can actually read geodatabases. However, we also have to know the layer name ahead of time.

First, let's read the layer that has the geometry in this gdb file.

```{r data, echo=TRUE, message = TRUE}
d.all <- sf::read_sf("./ACS_2020_5YR_COUNTY.gdb", layer = "ACS_2020_5YR_COUNTY")
glimpse(d.all)




#tmap::tm_shape(d.all) + tm_polygons() # commented out because 
#it's a large dataset that takes a long time to plot

```

Next, we're going to read the geodatabase again, but this time a tablular dataset containing age and sex data. Because the GEOID in this table and in the geometry file are different, we need to create a geoid in this table that will match. We *could* do it as a separate step, but we're also able to modify the table as we read it.
```{r read_tab, echo = T, message = TRUE}


d.x1 <- sf::read_sf("./ACS_2020_5YR_COUNTY.gdb", layer = "X01_AGE_AND_SEX") %>%
  mutate(fixed_geoid = str_sub(GEOID, start = 8, end = -1)) 
# the -1 is a shortcut to tell R to go to the end of the string

```

Finally, we're going to join the tabular and spatial data together. 

```{r join_data, echo = T, message = TRUE}

d.joined <- d.all %>% left_join(., d.x1, by = c("GEOID" = "fixed_geoid"))


```


Again, the data are too large, so we need to creat a subset we can work with later. Let's use the state ID to create a dataset with only those counties in Ohio. Be sure to check the data type of state ID

```{r make_subset, echo=TRUE, message = TRUE}

# get just Ohio
ohio <- d.joined %>% dplyr::filter(STATEFP == "39")


# map it to verify
tmap::tm_shape(ohio) + tm_polygons()

```

Next, we'll formalize our space by creating neighbors, and thus, **W**

- First we'll project
- Next, we'll use Queen contiguity to define **W**

```{r make some neighbors, echo=TRUE, message = TRUE}

# Check it first
sf::st_crs(ohio) 

# then reproject to north american equidistant conic
ohio.projected <- ohio %>% sf::st_transform(., "ESRI:102010")

# plot it again to make sure nothing broke
tmap::tm_shape(ohio.projected) + tm_polygons()

# make the neighborhood
nb <- spdep::poly2nb(ohio.projected, queen = TRUE)

nb

```

For each polygon in our polygon object, ```nb``` lists all neighboring polygons. For example, to see the neighbors for the first polygon in the object, type:

```{r checktheneighbors, echo = TRUE, message = TRUE}
nb[[1]]
```

Polygon 1 has seven neighbors. The numbers represent the polygon IDs as stored in the spatial object ```ohio.projected```. Polygon 1 is associated with the County attribute name `"Hancock County"` and its  neighboring polygons are associated with the counties:

```{r check names, echo = TRUE, message = TRUE}
ohio.projected$NAMELSAD[1] # county in index 1

nb[[1]] %>% ohio.projected$NAMELSAD[.] # and it's neighbors. 
# Note we're doing this programmatically step-by-step
```

Next, we need to assign weights to each neighboring polygon. In our case, each neighboring polygon will be assigned equal weight `(style="W")`. This is accomplished by assigning the fraction: `1 / ( # of neighbors)`  to each neighboring county then summing the weighted  values. While this is the most intuitive way to summaries the neighbors’ values it has one drawback in that polygons along the edges of the study area will base their lagged values on fewer polygons thus potentially over- or under-estimating the true nature of the spatial autocorrelation in the data. For this example, we’ll stick with the `style="W"` option for simplicity’s sake but note that other more robust options are available, notably `style="B"`.


```{r make weights, echo = TRUE, message = TRUE}
lw <- nb2listw(nb, style="W", zero.policy=TRUE)
```

The ```zero.policy=TRUE``` option allows for lists of non-neighbors. This should be used with caution since the user may not be aware of missing neighbors in their dataset. However, a zero.policy of `FALSE` would return an error if you have a dataset where a polygon does not have a neighbor.

To see the weight of the first polygon’s four neighbors type:

```{r checkit, echo = TRUE, message = TRUE}
lw$weights[1]
```

This row-normalized our weights!

We can also plot the distribution of neighbors across the dataset. 

```{r plotneighbors, echo = TRUE, message = TRUE}
# use attr to get the count of neighbors in W
neighbors <- attr(lw$weights,"comp")$d 

# plot it
hist(neighbors)

```


Finally, we’ll compute the average neighbor population of Females 75-79 years of age for each polygon. These values are often referred to as spatially lagged values. The following table shows the average neighboring F 75-79 values (stored in the F75.lag object) for each county. Note, I deterimined the correct attribute (B01001e47) by reading the metadata from the original Census Bureau link


```{r laggedVar, echo = TRUE, message = TRUE}
F75.lag <- lag.listw(lw, ohio.projected$B01001e47)
F75.lag
```

### Computing Moran's I


To get the Moran’s I value, simply use the moran.test function.

```{r morantest, echo = TRUE, message = TRUE}
moran.test(ohio.projected$B01001e47, lw)
```

Note that the p-value computed from the `moran.test` function is not computed from an MC simulation but **analytically** instead. This may not always prove to be the most accurate measure of significance. To test for significance using the MC simulation method instead, use the moran.mc function.


### Moran's plots

Thus far, our analysis has been a global investigation of spatial autocorrelation. We can also use local indicators of spatial autocorrelation (LISA) to analyze our dataset. One way of doing so is through the use of a Moran plot.

The process to make a plot is relatively simple:

```{r moranplot, echo = TRUE, message = TRUE}
# use zero.policy = T because some polygons don't have neighbors
moran.plot(ohio.projected$B01001e47, lw, zero.policy=TRUE, plot=TRUE)
```



## Your tasks

```{r}
#1. Create a spatial subset of the US, with at AT MINIMUM 4 states, MAXIMUM 7 states. States must be contiguous. Save this subset as a shapefile such that it's sufficiently small in size that GitHub will accept the git-push

# Define 5 contiguous states
my_states <- c("OH", "PA", "WV", "IN", "MI")

my_subset <- get_acs(
  geography = "county",
  state = my_states,
  variables = "B01003_001", # Total population (to pull the geometry)
  geometry = TRUE,
  cb = TRUE, # cb = TRUE will ensure the file is small enough for GitHub
  year = 2020
)

#Reproject to North America Equidistant Conic
my_subset_proj <- st_transform(my_subset, "ESRI:102010")

#Create a data folder and save the shapefile
dir.create("data", showWarnings = FALSE)
st_write(my_subset_proj, "data/my_spatial_subset.shp", delete_dsn = TRUE)

#Map
library(tmap)
tm_shape(my_subset_proj) + tm_polygons()
```


```{r}
#2. Choose a variable. If it's a raw count, you should normalize the variable in an appropriate manner (e.g., by total population, percent, by area)

# We will use Educational Attainment for 2020.
# B15003_001 = Total population 25 years and over
# B15003_022 = Population with a Bachelor's degree

edu_data <- get_acs(
  geography = "county",
  state = c("OH", "PA", "WV", "IN", "MI"),
  variables = c(total_pop = "B15003_001", bachelors = "B15003_022"),
  output = "wide",
  geometry = TRUE,
  cb = TRUE,
  year = 2020
)

# Normalize the raw count into a percentage
my_subset <- edu_data %>%

# Calculate the percentage of the population with a Bachelor's degree
  mutate(pct_bachelors = (bachelorsE / total_popE) * 100) %>%
  # Remove any counties with missing data to prevent errors later
  filter(!is.na(pct_bachelors)) %>%
  
# Reproject to North America Equidistant Conic 
  st_transform("ESRI:102010")

#see the first few rows of data table:
#head(my_subset)

#I like to see the data table pop in front of me. 
#View(my_subset)

#visualize task 2 
plot(my_subset["pct_bachelors"], main = "Percent Bachelor's Degrees")

```

```{r}
#3. Make a histogram of your chosen variable
hist(my_subset$pct_bachelors,
     main = "Distribution of Bachelor's Degrees",
     xlab = "Percentage of Population (%)",
     col = "lightblue",
     breaks = 20)
```

```{r}
#4. Make a choropleth map of your chosen variable. Choose an appropriate data classiﬁcation scheme

tm_shape(my_subset) +
  tm_polygons(
    col = "pct_bachelors",
    style = "jenks",
    palette = "Purples",
    title = "Population with a Bachelor's Degree (%)",
    border.alpha = 0.5,
    lwd = 0.5
  ) +
  tm_layout(
    legend.outside = TRUE, 
    legend.outside.position = "right", 
    frame = FALSE                        
  )
```


```{r}
#5. Develop a contiguity-based spatial weights matrix of your choosing (i.e., rook or queen)

#Creating a Queen contiguity neighborhood
nb_queen <- poly2nb(my_subset, queen = TRUE) 

# 5.1 Row-standardize the W
lw_queen <- nb2listw(nb_queen, style = "W", zero.policy = TRUE)

# 5.2 Plot a histogram of the number of neighbors
#use lengths() to count how many neighbors each county has
neighbor_counts <- lengths(nb_queen)

hist(neighbor_counts,
     main = "Histogram of Queen Neighbors",
     xlab = "Number of Neighbors",
     col = "Purple",
     breaks = seq(0, max(neighbor_counts) + 1, by = 1))

# 5.3 Calculate the average number of neighbors
avg_neighbors <- mean(neighbor_counts)
print(paste("Average number of neighbors:", round(avg_neighbors, 2)))


```
```{r}
# 5.4 Make a Moran Plot
moran.plot(my_subset$pct_bachelors, 
           listw = lw_queen, 
           zero.policy = TRUE,
           main = "Moran Plot: % Bachelor's Degrees",
           xlab = "County % Bachelor's",
           ylab = "Spatially Lagged % Bachelor's")
```


```{r}
#6. Repeat #5 (and 5.1 - 5.4) above with a W developed using the IDW method. You will need to investigate the `spdep` documentation to find the correct method/function.

# First, extract the center coordinates of each county. 
coords <- st_coordinates(st_centroid(st_geometry(my_subset)))

# Find a minimum distance threshold so every county gets at least 1 neighbor
k1 <- knn2nb(knearneigh(coords, k = 1))
max_dist <- max(unlist(nbdists(k1, coords)))

# Create the neighborhood list based on that distance
nb_dist <- dnearneigh(coords, d1 = 0, d2 = max_dist)

# Calculate the actual distances between the neighbors
dists <- nbdists(nb_dist, coords)

# Invert the distances for IDW (1 / distance)
idw_weights <- lapply(dists, function(x) 1 / x)

# 6.1 Row-standardize the weights matrix using the IDW
lw_idw <- nb2listw(nb_dist, glist = idw_weights, style = "W", zero.policy = TRUE)

# 6.2 Plot a histogram of the number of neighbors
hist(lengths(nb_dist), 
     main = "Histogram of Neighbors (Distance-based)", 
     xlab = "Number of Neighbors", 
     col = "lightblue")

# 6.3 Calculate the average number of neighbors
avg_idw_neighbors <- mean(lengths(nb_dist))
print(paste("Average number of IDW neighbors:", round(avg_idw_neighbors, 2)))

```
```{r}
# 6.4 Make a Moran Plot using the IDW weights
moran.plot(my_subset$pct_bachelors, 
           listw = lw_idw, 
           zero.policy = TRUE,
           main = "Moran Plot: % Bachelor's Degrees (IDW)",
           xlab = "County % Bachelor's",
           ylab = "Spatially Lagged % Bachelor's")
```

## Questions:

```{r}
#1. Describe in your own words how Moran’s I is calculated? 

#Answer: Moran’s I measures the degree of spatial autocorrelation, meaning whether similar values are located near each other. It compares the value at each location with the values of its neighboring locations. If nearby areas have similar values (high with high or low with low), Moran’s I is positive. If nearby areas are very different, it becomes negative.

#2. Describe in your own words: what is a spatially-lagged variable? 

#Answer: A spatially-lagged variable is the average (or weighted average) of values from neighboring areas for each location. It represents the local spatial context and shows what the surrounding areas look like compared to a given location.

#3. How does your analysis in this lab (as simple as it is) differ by how you have formalized W (e.g., space, neighbors) in two diﬀerent methods? How might it affect analysis? 

#Answer: The difference is how neighbors are defined. The Queen method defines neighbors based on shared borders, meaning any counties that touch are considered equal neighbors. In contrast, the IDW method defines neighbors based on distance, where closer counties have a stronger influence and farther counties have less influence. The effect is that Queen can sometimes be unrealistic, especially when large counties touch but are physically far apart. IDW is often more realistic because it considers actual distance, which better reflects how real-world processes (such as population patterns or ecological processes) change gradually over space.

#4. What does it mean if an observation falls in the “H-L” quadrant? Why might it be useful to detect such occurances? 

#Answer: "H-L" means a county has a High value but is surrounded by neighbors with Low values. This is useful for spotting spatial outliers or unique local anomalies (like a highly-educated college town surrounded by rural farmland) that don't match the broader regional trend.
```


## Bonus (+30 points)


```{r}
#B1. make another Moran plot, this time do so manually (use `geom_point` from `ggplot`). You must label each quadrant with HH, HL, LL, and LH, respectively. You should also use color and/or shape to denote whether an observation is statistically significant. Tip, you can find the data you want using the ```moran.plot``` function, but you'll have to alter the function call and read some documentation.

#Get the data
mp <- moran.plot(my_subset$pct_bachelors, listw = lw_queen, zero.policy = TRUE, plot = FALSE)

#ggplot
ggplot(mp, aes(x = x, y = wx)) +
  geom_point(aes(color = is_inf)) +
  geom_hline(yintercept = mean(mp$wx), linetype = "dashed") +
  geom_vline(xintercept = mean(mp$x), linetype = "dashed") +
  
  # Adding all 4 labels
  annotate("text", 
           x = c(max(mp$x), min(mp$x), max(mp$x), min(mp$x)), 
           y = c(max(mp$wx), min(mp$wx), min(mp$wx), max(mp$wx)), 
           label = c("HH", "LL", "HL", "LH"), fontface = "bold") +
  labs(title = "Moran Scatterplot: % Bachelor's Degrees", x = "County %", y = "Lagged %")
```


```{r}
#B2. plot a choropleth map of your dataset with a categorical color scheme, where the shading corresponds to the Moran plot (really, “LISA”) quadrants. Thus, your map will have four shades of color.

# assign HH, HL, LL, LH based on the averages
my_subset$quad <- ifelse(mp$x > mean(mp$x),
                         ifelse(mp$wx > mean(mp$wx), "High-High", "High-Low"),
                         ifelse(mp$wx > mean(mp$wx), "Low-High", "Low-Low"))

#plot the map
tm_shape(my_subset) +
  tm_polygons("quad", 
              palette = c("High-High"="red", "Low-Low"="blue", "High-Low"="pink", "Low-High"="lightblue"),
              title = "Clusters") +
  tm_layout(main.title = "Spatial Cluster Map: % Bachelor's Degrees")
```

